Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2005-04-11 paper/metamonads.tex
small corrections

The monad in this paper is similar to the one used in \cite{KiselyovTaha}.
However, the latter work used only |retN| of all monadic operations,
and used fixpoints (for performing iterations at generation time).
In this paper we do not use monadic fixpoints (because the generator
is not recursive) but we make extensive use of monadic operations for
generating conditional and looping operations.

|Blitz++| \cite{Veldhuizen:1998:ISCOPE} and in general template 
meta-programming in C++ achieve similar results of eliminating levels
of abstraction.  Using traits and concepts, some domain-specific
knowledge can also be encoded.  However inlining critically depends on
the compiler's fully inlining of all methods, and as has been reported
in the literature, this can be challenging to insure. Furthermore, all errors
(such as type errors and concept violation errors, that is,
composition errors) are detected only when compiling the generated
code. It is immensely difficult to correlate errors (e.g., line numbers) 
to the ones in the generator itself. Furthermore, as templates
cannot be local, a lot of code generation is scattered around instead
of being modularized.

@@

There are many more aspects which can also be handled
Input variations (augmented
matrices), error reporting (i.e. asking for the determinant of a 
non-square matrix), memory hierarchy issues, loop-unrolling
\cite{Padua:MetaOcaml:04},
warnings when zero-testing is undecidable and
a value is only probabilistically non-zero, etc.  The larger program
family of LU decompositions contains more aspects still.

@@

\section{Conclusion}\label{conclusion}
The combination of stateless modules (functors and structures), and our
monad with the compositional state makes aspects
freely composable without having to worry about the presence or absence of
value aliasing. The only constraints to compositionality are the
typing constraints plus the constraints we specifically impose,
including semantic constraints (e.g., rings do not have full
division).

>>>>

The monad in this paper is similar to the one used in \cite{KiselyovTaha}.
However, the latter work used only |retN| of all monadic operations,
and fixpoints (for performing iterations at generation time).
This paper does not involve monadic fixpoints (because the generator
is not recursive) makes extensive use of monadic operations for
generating conditional and looping operations.

|Blitz++| \cite{Veldhuizen:1998:ISCOPE} and in general template
meta-programming in C++ achieve similar results of eliminating levels
of abstraction.  Using traits and concepts, some domain-specific
knowledge can also be encoded.  However overhead elimination
critically depends on the compiler's fully inlining of all methods,
which has been reported to be challenging to insure. Furthermore, all
errors (such as type errors and concept violation errors, i.e.,
composition errors) are detected only when compiling the generated
code. It is immensely difficult to correlate errors (e.g., line
numbers) to the ones in the generator itself.

@@

There are many more aspects which can also be handled:
Input variations (augmented
matrices), error reporting (i.e. asking for the determinant of a 
non-square matrix), memory hierarchy issues, loop-unrolling
\cite{Padua:MetaOcaml:04},
warnings when zero-testing is undecidable and
a value is only probabilistically non-zero, etc.  The larger program
family of LU decompositions contains more aspects still.

@@

\section{Conclusion}\label{conclusion}
In this paper we demonstrate numerical code extensively parameterized
by complex aspects at no run-time overhead.  The combination of
stateless functors and structures, and our monad with the
compositional state makes aspects freely composable without having to
worry about value aliasing. The only constraints to compositionality
are the typing constraints plus the constraints we specifically
impose, including semantic constraints (e.g., rings do not have full
division).
